\chapter{Introduction}
Welcome to the Software Craftsmanship Javascript Workbook. This workbook will
lead you through a variety of programming exercises in conjunction with the main
Software Craftsmanship book. Programming is a skill, and like any other skill,
it takes practice to become truly profficient. The exercises in this book range
from ``monkey-see-monkey-do" problems getting you used to typing exactly as a
computer expects, to free-ranging ideas on programs you might be interested in
writing for yourself. 

You will want to complete every ``MSMD" section exxctly as written. Computers
are not forgiving when it comes to what you type in your program, and these
exercises are a safe way to begin exploring the world of computer programming.
The exercises for each chapter build on the MSMD section, and are opportunities
for you to apply what you've learned conceptually to a program. You should make
an attempt to complete all the exercises. Finally, projects are larger scale
ideas for programs that you would be able to write with the tools presented by
that point in the main text. The more you program, the better a programmer
you'll be, but you'll probably want to pick and choose projects that interest
you.

\section{node.js}
Javascript as a programming language started as a tool to make web pages
dynamic. While the web browser is still the most common platform for Javascript
to run (by an incredibly large margin), it is becoming more popular as a
general-purpose language. Still, it's not installed by default on many
computers, so we'll be installing a program called node.js to run your
Javascript.

\section{First Program}
Let's write our first program. Open your text editor (gedit), and type the 

\subsection{hello.js}
\begin{lstlisting}
console.log("Hello" + ", " + "world!");
console.log(2 + 3, 10/7, 152 * 12.6);
\end{lstlisting}

\subsection{Running the program}
Open your console. Type {\tt node hello.js}

\begin{lstlisting}
Hello, world!
5 1.4285714285714286 1915.2
\end{lstlisting}

Congratulations! You've written your first program!
